\subsection{Age of Information foundations}
The Age of Information (AoI) was introduced in \cite{kaul2012aoi} as a
destination-centric measure of data freshness, capturing the time elapsed
since the generation of the most recently received update. The concept has
since matured into a substantial body of theory, surveyed comprehensively in
\cite{yates2021aoi} and developed as a self-contained metric and analysis tool
in \cite{kosta2017aoi}. Two threads of this literature are directly relevant to
our problem. The first concerns sampling and transmission policy: \cite{sun2017update}
shows that, contrary to intuition, freshness is not always improved by updating
as often as possible, and that an optimal policy may deliberately wait, while
\cite{costa2016age} analyzes how buffer and packet management shape the age
process. The second concerns scheduling across contending sources:
\cite{kadota2018scheduling} derives age-optimal and near-optimal policies for
broadcasting updates from many sources over a shared wireless channel. These
foundations motivate our use of peak AoI as the design objective and, more
subtly, they anticipate the tension our formulation confronts, namely that
serving more sources with a shared resource (here, a shared charger) can raise
age once contention is taken into account.

\subsection{UAV communications, trajectory, and placement}
Unmanned aerial vehicles have become a central platform for flexible wireless
service, with the opportunities and open problems laid out in \cite{zeng2016uav}
and a broad tutorial treatment in \cite{mozaffari2019tutorial}; the networking
of cooperating aerial nodes is surveyed in \cite{bekmezci2013fanet}. Trajectory
design is the dominant control lever in this setting: \cite{wu2018multiuav}
jointly optimizes the paths and communication of multiple UAVs, and for data
collection specifically \cite{zhan2018energy} minimizes collection energy while
\cite{gong2018flight} minimizes the flight time needed to gather data from a
field of sensors. The quality of the air-to-ground link that these designs
exploit is governed by the elevation-dependent line-of-sight probability
characterized in \cite{alhourani2014optimal} and by the aerial channel models
standardized in \cite{tr36777}, which we draw on when specifying the collection
link. Two combinatorial structures recur in our own design. Deciding where to
site a small number of charging stations is a facility-location problem, whose
models and algorithms are treated in \cite{daskin2013}. Planning a visiting tour
in which a source is served whenever the UAV enters its communication range is a
close-enough traveling salesman problem, formulated and solved exactly in
\cite{behdani2014cetsp} and heuristically in \cite{cetsp}. Our formulation
couples these two structures, which are typically studied in isolation, to the
freshness and charging dynamics described below.

\subsection{AoI-oriented UAV data collection}
Building on these foundations, AoI has become a standard objective for
UAV-assisted data collection, where a UAV visits distributed sources and the
design goal is to keep the collected information fresh. Single-UAV formulations
establish the basic tradeoffs: \cite{liu2021uav} designs UAV-aided collection to
maximize information freshness across a sensor field, and \cite{abdelmagid2019average}
minimizes average peak AoI in a UAV-assisted IoT network, jointly treating
trajectory and update scheduling. In the swarm setting, \cite{eldeeb2023swarm}
minimizes AoI across multiple cooperating UAVs using multi-agent reinforcement
learning, demonstrating that coordinated trajectories can reduce peak AoI
relative to uncoordinated flight; learning-based control of multiple UAVs is
itself an active area, with resource allocation studied in \cite{cui2020multiagent}
and deployment and movement design in \cite{liu2019rl}. That family of
formulations, however, assumes flight endurance is unconstrained: it carries no
energy state and no recharging model, so the depletion and replenishment
dynamics that dominate long missions are absent. Closest to our objective is
\cite{base2024}, which jointly optimizes charging station placement, UAV
trajectory, and peak AoI, and which we adopt as our point of departure. That
work is nonetheless single-UAV: with one vehicle there is no inter-UAV
contention for a charger, no need for multi-port capacity, and no swarm
assignment, precisely the coupling that our design targets.

\subsection{Energy-limited UAVs and charging}
A second line of work makes energy replenishment explicit. The battery lifetime
of aerial nodes is a first-order operational constraint, as argued in
\cite{galkin2019uavs}, and wireless power transfer offers one replenishment
mechanism whose trajectory and energy tradeoffs are optimized in \cite{xu2018wpt}.
Turning to shared charging infrastructure, \cite{wei2024} lets multiple UAVs
share a limited charging facility and folds the resulting charge waiting time
into the AoI objective, then optimizes it by deep reinforcement learning. The
facility there is single and fixed in location, which effectively reduces
contention to a single-server queue and leaves station placement out of scope.
Laser-based (LBD) charging is studied in \cite{sun2025laser}, where charging
sites are at fixed locations and are treated without queue contention, so
neither where to place stations nor how many vehicles may wait is optimized. A
different tactic uses mobile chargers: \cite{oubbati2024ugv} and \cite{zhao2025aoi}
employ ground vehicles that follow the UAVs to deliver energy on the move.
Because the charger comes to the UAV, this removes the fixed-station contention
that is central to our setting and does not address the placement of fixed,
multi-port stations. Finally, \cite{ev2022sharing} does model UAV charging
stations as a contended queue and provides a waiting-time analysis, but it
optimizes neither AoI nor trajectory nor station placement, so the queueing
model is not coupled to any freshness or mobility decision. When we make
contention explicit, we model a station as a finite-source (closed) service
system rather than an open queue. This places our analysis in the lineage of the
machine-repair, or machine-interference, problem surveyed in \cite{haque2007machine}
and treated within the classical queueing framework of \cite{grossharris}, from
which we adapt the finite-source multi-server results used in Section~\ref{sec:model}.

\subsection{Positioning and gap}
Read together, these works each supply a subset of the ingredients our problem
requires but never their combination. Swarm operation with AoI appears in
\cite{eldeeb2023swarm} yet without charging; placement with AoI appears in
\cite{base2024} yet only for a single UAV; charge waiting is captured in
\cite{wei2024} yet at a single fixed server without placement; and a contended
charging queue is analyzed in \cite{ev2022sharing} yet without AoI,
trajectory, or placement. No prior formulation simultaneously carries all four
of station placement, swarm operation, multi-server charging contention, and a
peak-AoI objective. Our work closes this gap by jointly optimizing station
placement, port capacity, swarm trajectory, and source-to-UAV assignment under
a charging model that is a contended, multi-server, and finite-source (closed)
queue, validated against discrete-event simulation. The finite-source aspect
is not a cosmetic refinement: because the number of UAVs competing for a
station is small and closed, modeling the queue as an open M/M/c system
over-predicts waiting time and thereby distorts both placement and capacity
decisions, an effect we quantify and correct with the finite-source model
detailed in Section~\ref{sec:model}. This coupled formulation also yields two
structural findings that a decoupled treatment cannot expose: a congestion
threshold, beyond which peak AoI is non-monotone in swarm size, and a
placement inversion, whereby coverage-optimal station placement becomes
AoI-suboptimal once contention is accounted for.
