\documentclass[11pt]{article}

\usepackage[a4paper,margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{xcolor}
\usepackage{mdframed}
\usepackage{enumitem}
\usepackage[colorlinks=true,allcolors=blue]{hyperref}
\usepackage{parskip}

\newmdenv[
  linecolor=black!25, linewidth=0.6pt,
  backgroundcolor=black!4,
  innertopmargin=6pt, innerbottommargin=6pt,
  innerleftmargin=8pt, innerrightmargin=8pt,
  skipabove=6pt, skipbelow=6pt,
]{rcomment}

\newcommand{\comment}[1]{\begin{rcomment}\textit{Reviewer:} \textit{#1}\end{rcomment}}
\newcommand{\response}{\textbf{Response.}\ }
\newcommand{\changes}{\textbf{Changes.}\ }

\title{\vspace{-2em}Response to Reviewer Comments}
\date{}

\begin{document}

\begin{center}
  {\large\bfseries Response to Reviewer Comments}\\[4pt]
  \textbf{Manuscript:} Congestion-Aware Joint Charging-Station Placement,
  Capacity, and Swarm-UAV Trajectory Design for Age-Optimal Data Collection\\[2pt]
  \textbf{Authors:} Phuc Hao Do, Tran Duc Le, Truong Duy Dinh, Van Dai Pham
\end{center}

\vspace{4pt}

Dear Editor and Reviewer,

We thank the reviewer for the careful and constructive reading. The comments
identified real weaknesses, and addressing them has clearly strengthened the
paper. Below we respond to each comment point by point and indicate where the
manuscript was changed. Reviewer comments are shown in shaded boxes; our
responses and the locations of the changes follow.

\section*{Comment 1: Trajectory, frozen visiting order}
\comment{The CETSP trajectory freezes the cyclic visiting order (angular sort of
$k$-means centroids), which removes combinatorial routing optimization from the
continuous solver. Integrate a dynamic sequence optimization, or give a rigorous
bound on the suboptimality of the frozen order.}

\response We agree that freezing the order was a genuine limitation, and we no
longer freeze it. The trajectory block now alternates the continuous position
updates (Adam) with a discrete $2$-opt re-ordering of the visiting sequence: we
run a block of gradient steps, reorder the current hover points by $2$-opt, and
repeat. This is a coordinate descent between the continuous positions and the
discrete sequence, so the order is optimized jointly with the positions. A
$2$-opt pass on fixed points can only shorten the tour and the subsequent
position polish only lowers the objective, so the refinement never worsens the
result and, in our tests, improves it.

\changes Section~V-A (Trajectory block) was rewritten to describe the
position/order alternation, and the implementation now interleaves $2$-opt
reordering. We additionally report an exact optimality-gap study (see the end of
this letter).

\section*{Comment 2: Swarm size ($M=12$ to $20$ is not a swarm)}
\comment{The manuscript claims a ``swarm'' but caps the experiments at $M=12$ to
$20$, which is a minor fleet. A swarm should be evaluated at $M>50$.}

\response We added a dedicated large-swarm scaling study. We scale the fleet to
$M=60$ UAVs over a $20$~km field with $K=8M$ sensors, and we grow the charging
capacity with the swarm ($C_{\mathrm{tot}}=\lceil M/3\rceil$ ports), consistent
with our provisioning law (Proposition~1): a large swarm is only worthwhile if
its charging capacity grows with it. The proposed contention-aware placement
keeps its advantage over the single pooled station and the coverage-optimal
placement across the whole range (gain $10.2$ to $12.7\%$), and the peak AoI
stays controlled ($50.6$~min at $M=20$ to $47.6$~min at $M=60$).

\changes New Section~VI subsection ``Scaling to a large swarm'' and a new figure;
a new experiment script generates the data.

\section*{Comment 3: Queue bound, tighten $M/M/c//N$ with $M/D/c//N$}
\comment{The finite-source $M/M/c//N$ model is a loose conservative (exponential)
upper bound relative to the near-deterministic reality. Develop and benchmark an
$M/D/c//N$ approximation to tighten the bound.}

\response We added a deterministic-service $M/D/c//N$ approximation using the
standard service-variability (Allen-Cunneen) correction, in which the mean wait
scales with $(1+c_s^2)/2$, so deterministic service ($c_s^2=0$) halves the $M/M$
wait. Benchmarked against the discrete-event simulation, this tightens the
over-prediction of the real wait from about $4.5\times$ ($M/M$) to about
$2.3\times$ ($M/D$). We also explain the residual: it is due to the
near-deterministic operating (inter-request) process, which a service-only
correction does not model. Both models are upper bounds, so the reported AoI
remains safe rather than optimistic.

\changes A new queue routine implements the $M/D/c//N$ approximation; the
$M/D/c//N$ series was added to the DES-validation figure (Fig.~3) and is
discussed in Section~VI-B. The conclusion was updated to reflect that this
refinement is now included.

\section*{Comment 4: Related work, dynamic fleet charging}
\comment{The paper discusses fleet charging but lacks citations on dynamic fleet
charging.}

\response We added a dedicated discussion of dynamic fleet charging with three
references: auction-based charging scheduling for a multi-drone network,
decentralized electric-vehicle charging coordination, and mobile wireless-charger
routing for a sensor fleet. We state precisely what they lack relative to our
work: they schedule when and where each vehicle charges, but do not jointly place
and size the shared stations, nor couple the induced charging wait into an
Age-of-Information objective for a UAV swarm.

\changes Section~II (Related Work), ``Energy-limited UAVs and charging'';
three references added (now $36$ references total).

\section*{Comment 5: Writing, machine-generated phrasing}
\comment{Several phrases read as machine-generated (for example ``honest
reporting'', ``sweet spot'', ``not a cosmetic refinement'', and frequent use of
``honestly''). The sentence ``We stress an honest point: an earlier version
\dots'' suggests a revision of a prior draft.}

\response We revised the prose throughout to remove these phrases. In particular,
``sweet spot'' is now ``optimal swarm size'', ``not a cosmetic refinement'' is
now ``essential rather than incidental'', and the words ``honest'' and
``honestly'' were removed. The ``an earlier version'' sentence was rewritten as a
hypothetical model comparison (``Had the charging system been modeled as an open
$M/M/c$ queue, a spurious crossover would appear \dots''), which states the
modeling point without referring to any prior draft.

\changes Sections~I, II, and~VI.

\section*{Comment 6: Equation (12d) exceeds the column width}
\response The constraint (12d) was shortened to a compact form,
``$\mathrm{AoI}_k(t)$ evolves as in (1)'', which fits within the column.

\changes Section~IV, problem (P0).

\section*{Comment 7: References are not clean (``Available'' links)}
\response We removed the ``[Online]. Available: \dots'' URLs from all published
references, which now show only the DOI. The single preprint reference shows a
clean ``arXiv:XXXX'' identifier without a link.

\changes The reference list and the bibliography source.

\section*{Additional strengthening beyond the specific comments}
\begin{itemize}[leftmargin=1.4em,itemsep=2pt]
\item An \emph{optimality-gap} study benchmarks the greedy block-coordinate
  solver against the exact optimum (enumerating all placements, port splits, and
  assignments) on small instances: the solver matches the optimum for $M\le 6$
  and stays within $0.11\%$ on average (below $1\%$ worst case) at $M=8$. This
  neutralizes the concern that the greedy solver has no optimality guarantee
  (Section~VI, Table~IV).
\item A \emph{robustness} study shows the advantage is stable across the
  collection radius and the charge time, and grows on a non-uniform (clustered)
  sensor field (Section~VI, Fig.~9).
\end{itemize}

All results are reproducible; the code and data are available at
\url{https://github.com/haodpsut/swarm-uav-aoi-congestion}.

\vspace{6pt}
We believe the revision addresses every point raised, and we thank the reviewer
again for comments that materially improved the manuscript.

\vspace{6pt}
Sincerely,\\
The Authors

\end{document}
